# Observable

Compared with Shiny and Dash, Observable seems like another world:

 - It is generally used as a hosted service.
 
 - Virtually everything runs in the user's browser.
 
 - Reactivity is baked in to everything.
 
 - It uses (a very close approximation to) JavaScript.
 
That said, there's a few things from your R and tidyverse world that may help you get acquainted:

 - Functional-programming ideas translate well from R to JavaScript.
 
 - There are a couple of "dplyr/tidyr"-like packages in JavaScript: arquero and tidyjs.
 
 - There are a couple of JavaScript visualization packages that use grammar-of-graphics: Vega-Lite and Plot.
 
I try to keep in mind that the point of the exercise, largely, is to "do stuff to data frames".
Knowing how to "do stuff" and "think about stuff" using tidyverse makes it easier for me to figure out the same "stuff" elsewhere.

## Principles

### Hosted service runs in browser

### Reactivity baked in

### JavaScript

### Tidyverse thinking helps

## Demonstration app

```{r echo=FALSE, fig.cap="Reactivity diagram"}
knitr::include_graphics("images/observable-aggregate-local.svg")
```

```{r echo=FALSE, fig.cap="Legend"}
knitr::include_graphics("images/observable-legend.svg")
```

### Showcase

### Workshop

### Appendix
